---
title: "AVR"
author: "Josh Havstad"
date: "2024-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Archetype Value Rating

## Overview

In this file, I hope to create the equation for my metric. This will rate players based on how they perform in specific roles (or archetypes as I will call it). If you are familiar with the game NBA 2k, then you know this is a term which is used to categorize or describe a player's playstyle. For example, a "playmaker" archetype can be used to describe players such as Chris Paul, Draymond Green, or Nikola Jokic. Playmakers can handle the ball, facilitate for others, and create good looks for other players. As you can see, this term can be used to describe players from different positions. Obviously, these players can be described with other archetypes or play-styles as well, but this is just a basic example. The reason I believe that rating players based off "archetypes" is a valuable metric is due to the new era of "positionless" basketball. Nowadays, you don't typically see your stereotypical true point guard or back-to-the-basket center. Positions are often just a technicality. In my opinion, lineups are mostly built around player "archetypes." An example of this would be the Celtics. Their forwards play the role of play-makers and shot creators, while their guards are defenders and off-ball shooters/cutters. Their big men can stretch the floor while also providing rim protection. This goes away from the older era of basketball and the duties you would normally associate with each position.

Creating a metric to rate players on the value they provide in specific archetypes or roles could create opportunities for:

-   A fun discourse on players' abilities to succeed in different roles

-   An evaluation on role players' contributions to their team

-   A ranking of players based on their success in their given role rather than overall talent

The last point is the idea I find really interesting and is the main motivation for this metric. As an NBA fan, I always wonder about a player's true value to their team. Is an inefficient number one option on a bad team truly more valuable than a role player who is excelling in his role and given opportunities? These role players I am referring to are the players that turn contenders into champions. Perhaps this metric could be used for teams to identify players who fulfill a role they need.

## How should archetypes be identified?

While I could go through each player and assign them to an archetype individually, this would be very subjective and based on my knowledge of each player.

A better approach would be a clustering method which can take into account all of the statistics in our data set and assign each player to a different cluster. Ideally, the clusters would reveal the "archetypes" themselves through the players that are in each one.

```{r,include=F}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(cluster)
library(rvest)
```

```{r}
NBA23 <- read.csv("~/Desktop/Basketball-Analytics/NBA23.csv", check.names=F)
head(NBA23)
```

```{r,results = 'hide'}
url <- "https://www.basketball-reference.com/leagues/NBA_2023_per_poss.html"
wp <- read_html(url)
per_poss <- html_table(html_nodes(wp, "#per_poss_stats"), fill = TRUE)[[1]]
write.csv(per_poss, "per_poss.csv", row.names = FALSE)
per_poss <- per_poss %>% select(Player, ORtg, DRtg)
per_poss <- per_poss[!duplicated(per_poss$Player),]
head(per_poss,10)
```

```{r,results = 'hide'}
NBA23 <- merge(NBA23, per_poss, by='Player')
head(NBA23)
```

```{r,results='hide'}
NBA23 <- NBA23 %>% mutate('GS%' = GS/G)
NBA23$ORtg <- as.numeric(NBA23$ORtg)
NBA23$DRtg <- as.numeric(NBA23$DRtg)
NBA23$Pos <- as.factor(NBA23$Pos)
head(NBA23)
```

```{r,warning=F}
NBA23 <- na.omit(NBA23)
NBA23pg <- mutate_at(NBA23, vars(c(8:10,12:13,15:16,19:20,22:30,42:44,49,64)), funs(./G))
names(NBA23pg)
```

We will use our NBA23 data set to cluster the players based on the various statistics. I added offensive rating, defensive rating, and games started percentage (games started/games played), as I thought they could be useful predictors during this process. I also changed many of the statistics from totals to per game metrics by dividing certain stats by games played. All of the statistics relating to the players are shown above.

### K-Means Clustering

```{r}
# select relevant variables for clustering
vars <- NBA23pg[,c(4,8:69)]

# Perform K-means clustering
k <- 9
set.seed(23) # for reproducibility
kmeans_result <- kmeans(vars, centers = k)

# Add cluster labels to the original dataset
NBA23pg$cluster <- as.factor(kmeans_result$cluster)

# Analyze the resulting clusters
cluster_summary <- NBA23pg %>% 
  group_by(cluster) %>% 
  summarise(mean_points = mean(PTS),
            mean_rebounds = mean(TRB),
            mean_assists = mean(AST),
            mean_steals = mean(STL),
            mean_blocks = mean(BLK))

print(cluster_summary)
```

```{r}
NBA23pg %>% filter(cluster==1)
NBA23pg %>% filter(cluster==2)
NBA23pg %>% filter(cluster == 3)
NBA23pg %>% filter(cluster == 4)
NBA23pg %>% filter(cluster == 5)
NBA23pg %>% filter(cluster == 6)
NBA23pg %>% filter(cluster == 7)
NBA23pg %>% filter(cluster == 8)
NBA23pg %>% filter(cluster == 9)
```
