---
title: "Archetype Value Rating"
author: "Josh Havstad"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(rvest)
```

# The Need for a New Statistic

## Overview

In this file, I will look into current, popular metrics which are used to evaluate players' performance. I will try to find if these metrics have flaws or natural biases in them. I will also try to find areas or aspects of basketball that are constantly undervalued or not valued at all. This will allow for a discussion on what the problems are with current metrics and how a new metric can potentially solve this issue.

## Problems with Current Metrics

To support the idea of a need for a new metric, I will investigate the current popular metrics and statistics used to evaluate players. Finding the faults or flaws with these will identify the areas that I need to focus on when creating my new player rating system.

### PER

First, let's look at the most popular statistic, PER. I am unsure on how teams, coaches, and players view this since I have never been involved in a professional organization, but this metric is referenced the most by the fans and the media when discussing a player's overall value for a given season.

PER was created by John Hollinger, and it "sums up all a player's positive accomplishments, subtracts the negative accomplishments, and returns per-minute rating of a player's performance." At the time it was created, PER was advanced due to the fact that it adjusted for pace of play and playing time, and it normalized data across seasons so that the average PER is always 15. Some criticisms of this metric, however, are that it overvalues bigs, undervalues defensive impact, and does not incorporate any advanced statistics.

```{r}
url3 <- "https://www.basketball-reference.com/leagues/NBA_2023_advanced.html"
wp3 <- read_html(url3)
advanced <- html_table(html_nodes(wp3, "#advanced_stats"), fill = TRUE)[[1]]
write.csv(advanced,"advanced.csv", row.names = FALSE)
advanced$PER <- as.numeric(advanced$PER)
head(advanced)
```

```{r}
blank <- which(is.na(advanced$PER))
advanced <- advanced[-blank,]
advanced <- advanced[,-c(20,25)]
```

```{r}
advanced %>%
  group_by(Pos) %>%
  filter(n()>2) %>%
  summarise(mean=mean(PER)) %>%
  arrange(desc(mean))
```

We can see from this table that centers and power forwards have the highest average PER values, supporting the idea that this rating favors big men.

```{r}
advanced$Pos <- as.factor(advanced$Pos)
advanced$Pos <- relevel(advanced$Pos, ref = "SF")
PERmodel <- lm(PER~Pos, data = advanced)
summary(PERmodel)
anova(PERmodel)
```

In the model above, I set small forward to the reference category since it had the lowest average PER. When running a linear model, we can see that the only two positions with a significant association with PER are Center and Power Forward. The coefficient estimate of Center tells us that we can expect the PER for a Center to be 4.3573 points higher than the PER for a Small Forward, on average.

Additionally, after running an Analysis of Variance, we see that position (p-value = 1.762e-06) is significantly associated to PER. This tells us that there are natural biases within the formulation of PER which are overvaluing certain aspects of basketball while simultaneously undervaluing other aspects of basketball. This is especially true when you consider the overwhelming amount of talent at the guard and wing positions which would make you think that those positions would have higher PER values than they do.

Obviously, nothing can be perfect, but the main flaw with PER is that it lacks the use of advanced statistics in its calculation. It is essentially a weighted combination of "box-score" statistics which every casual fan is familiar with. PER also tends to favor efficient players who do not turn the ball over as much, which in turn favors big men who shoot closer to the basket and typically do not have many ball-handling duties. However, I believe "efficiency" should be relative to a player's style of play and responsibilities on the floor while playing.

## Which statistics are associated with each position?

```{r}
library(cluster)
library(factoextra)
NBA23 <- read.csv("~/Desktop/Basketball-Analytics/NBA23.csv",check.names=F)
head(NBA23)
```

I will now use the "NBA23" data set I created in the Data Collection file to analyze which variables/statistics are closely related to each position in basketball. In turn, this will allow us to possibly identify which variables are being overestimated/underestimated when it comes to PER.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'PG'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
PG <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Point Guard\n")
print(PG)
```

The variables most significantly associated with the point guard position are assist percentage, percentage of their made 2 and 3 pointers which were assisted, number of assists, and total rebound percentage.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'SG'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
SG <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Shooting Guard\n")
print(SG)
```

The variables most significantly associated with the shooting guard position have to do with rebounding percentages, block percentage, 2 point percentage, and win shares.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'SF'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
SF <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Small Forward\n")
print(SF)
```

The variables most associated with the small forward position have to do with shot distances, assist percentage, turnover percentage, and percent of 2 point field goals assisted.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'PF'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
PF <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Power Forward\n")
print(PF)
```

The variables most significantly associated with the power forward position are rebounding percentages, percent of made shots that are assisted, dunks, and 2 point field goal percentage.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'C'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
C <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Center\n")
print(C)
```

Similar to the power forward position, the variables most significantly associated with the center position are rebounding percentages and dunks. Other important variables are block percentage, field goal percentage, and shot distance.

```{r}
PFC <- intersect(PF$Variable,C$Variable)
length(PFC)
PFC
```

We see that there are 20 variables which are significantly associated with both power forward and center.

```{r}
setdiff(PFC,PG$Variable)
setdiff(PFC,SG$Variable)
setdiff(PFC,SF$Variable)

```

## My Proposed Solution

...

## 
