---
title: "Archetype Value Rating"
author: "Josh Havstad"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(rvest)
```

# The Need for a New Statistic

## Overview

In this file, I will look into current, popular metrics which are used to evaluate players' performance. I will try to find if these metrics have flaws or natural biases in them. I will also try to find areas or aspects of basketball that are constantly undervalued or not valued at all. This will allow for a discussion on what the problems are with current metrics and how a new metric can potentially solve this issue.

## Problems with Current Metrics

To support the idea of a need for a new metric, I will investigate the current popular metrics and statistics used to evaluate players. Finding the faults or flaws with these will identify the areas that I need to focus on when creating my new player rating system.

### PER

First, let's look at the most popular statistic, PER. I am unsure on how teams, coaches, and players view this since I have never been involved in a professional organization, but this metric is referenced the most by the fans and the media when discussing a player's overall value for a given season.

PER was created by John Hollinger, and it "sums up all a player's positive accomplishments, subtracts the negative accomplishments, and returns per-minute rating of a player's performance." At the time it was created, PER was advanced due to the fact that it adjusted for pace of play and playing time, and it normalized data across seasons so that the average PER is always 15. Some criticisms of this metric, however, are that it overvalues bigs, undervalues defensive impact, and does not incorporate any advanced statistics.

```{r}
url3 <- "https://www.basketball-reference.com/leagues/NBA_2023_advanced.html"
wp3 <- read_html(url3)
advanced <- html_table(html_nodes(wp3, "#advanced_stats"), fill = TRUE)[[1]]
write.csv(advanced,"advanced.csv", row.names = FALSE)
advanced$PER <- as.numeric(advanced$PER)
head(advanced)
```

```{r}
blank <- which(is.na(advanced$PER))
advanced <- advanced[-blank,]
advanced <- advanced[,-c(20,25)]
```

```{r}
advanced %>%
  group_by(Pos) %>%
  filter(n()>2) %>%
  summarise(mean=mean(PER)) %>%
  arrange(desc(mean))
```

We can see from this table that centers and power forwards have the highest average PER values, supporting the idea that this rating favors big men.

```{r}
advanced$Pos <- as.factor(advanced$Pos)
advanced$Pos <- relevel(advanced$Pos, ref = "SF")
PERmodel <- lm(PER~Pos, data = advanced)
summary(PERmodel)
anova(PERmodel)
```

In the model above, I set small forward to the reference category since it had the lowest average PER. When running a linear model, we can see that the only two positions with a significant association with PER are Center and Power Forward. The coefficient estimate of Center tells us that we can expect the PER for a Center to be 4.3573 points higher than the PER for a Small Forward, on average.

Additionally, after running an Analysis of Variance, we see that position (p-value = 1.762e-06) is significantly associated to PER. This tells us that there are natural biases within the formulation of PER which are overvaluing certain aspects of basketball while simultaneously undervaluing other aspects of basketball. This is especially true when you consider the overwhelming amount of talent at the guard and wing positions which would make you think that those positions would have higher PER values than they do.

Obviously, nothing can be perfect, but the main flaw with PER is that it lacks the use of advanced statistics in its calculation. It is essentially a weighted combination of "box-score" statistics which every casual fan is familiar with. PER also tends to favor efficient players who do not turn the ball over as much, which in turn favors big men who shoot closer to the basket and typically do not have many ball-handling duties. However, I believe "efficiency" should be relative to a player's style of play and responsibilities on the floor while playing.

### WS/48

WS/48 is "Win shares per 48 minutes" and is a per-minute version of the popular metric, win shares, which measures a player's contribution to winning. Winning is the name of the game, and you would expect WS/48 to accurately rank which players are the best or have the most "impact" on success.

```{r}
NBA23 %>% arrange(desc(WS/48)) %>% select(Player) %>% head(25)
```

Above, we can see the players who led the league in WS/48 for the 2022-2023 season. Some notable names in the top 10 are Domantas Sabonis at number 2 and Jarrett Allen at number 8. We also see Evan Mobley, Nikola Vucevic, and Brook Lopez in the top 20.

```{r}
ws48 <- NBA23 %>% group_by(Pos) %>% filter(n() >2) %>% summarise(mean=mean(WS/48)) %>% arrange(desc(mean))

ggplot(ws48, aes(x = Pos, y = mean)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Average Win Shares by Position", x = "Position", y = "Average Win Shares per 48")
```

Similar to PER, we see that center and power forward have the best average rating when it comes to WS/48.

```{r}
NBA23$Pos <- as.factor(NBA23$Pos)
NBA23$Pos <- relevel(NBA23$Pos, ref = "SG")
WS48model <- lm(WS/48~Pos, data = NBA23)
summary(WS48model)
anova(WS48model)
```

From the model above which implemented an analysis of variance, we see that position has a significant association on WS/48. When comparing to the baseline category of shooting guard, we see that centers and power forwards have strong evidence of a significant difference in WS/48, and point guards have moderate evidence of a difference in WS/48.

### VORP

VORP is another popular advanced metric which stands for Value Over Replacement Player. As it says in the title, this statistic is meant to quantify a player's impact on their team compared to a hypothetical "replacement-level" player. A "replacement-level" player is essentially a player that is readily available and could be signed for a minimum-salary. VORP is calculated with various box-score statistics as well. Teams should aim for all their players to be better than some hypothetical minimum-salary player if they want to be competitive and successful. VORP values can be both positive and negative, with larger values indicating that a player contributes more to their team.

```{r}
# top 20 NBA players in VORP for 2022-2023 season
NBA23 %>% select(c(Player,VORP)) %>% arrange(desc(VORP)) %>% head(20)
```

Looking at the best rated players in VORP, we see that Nikola Jokic, Luka Doncic, and Joel Embiid are the only players over 6. Jokic clearly comes out on top with 8.8. These results make sense, as these players had some of the best seasons statistically speaking, and their teams would not be nearly as successful without them.

```{r}
# players who were below-average in VORP
NBA23 %>% select(c(Player,VORP)) %>% filter(VORP < mean(VORP)) %>% arrange(desc(VORP)) %>% head(20)
```

Looking at the players that finished the season with below-average VORP, we find surprising players such as Kentavious Caldwell-Pope, Jaden McDaniels, and Marcus Smart. I say this is surprising due to the fact that these players were instrumental in their teams' success during the 2022-2023 NBA season, as Caldwell-Pope started for the NBA champion Denver Nuggets. These players also served as the primary defenders against many of the top scoring talents in the league, and their team's success would often come down to if they were successful in slowing down the primary scorer on the other team.

```{r}
# box score stats for the 3 players
NBA23 %>% 
  filter(Player %in% c("Kentavious Caldwell-Pope","Jaden McDaniels","Marcus Smart")) %>%
  mutate_at(vars(PTS, TRB, AST, STL, BLK, TOV, MP), ~./G) %>%
  select(c(Player, Pos, PTS, 'FG%', TRB, AST, STL, BLK, TOV, MP))
```

From the table above with the box-score statistics, we see that these players were not huge contributors in counting stats despite all averaging over 30 minutes a night, with none of the players averaging more than 13 points or 4 rebounds. This could explain their low VORP rating, as the statistic factors in these statistics.

```{r}
# players who were above-average in VORP
NBA23 %>% select(c(Player,VORP)) %>% filter(VORP > mean(VORP)) %>% arrange(VORP) %>% head(20)
```

On the flip side, we can see players who finished slightly above-average in VORP (and better than Kentavious Caldwell-Pope, Jaden McDaniels, and Marcus Smart) in the table above. Some players I found surprising were Robert Covington, Thomas Bryant, Jordan Goodwin, and Kenrich Williams. This is not meant to diminish these players, but I would not consider any of those players more impactful or better than the three players I previously looked into that were below-average in VORP. These four players who were above-average in VORP struggled to find consistent minutes on the teams they played for because they simply are not as talented as some of their counterparts in the NBA.

```{r}
# box score stats for the 4 players
NBA23 %>% 
  filter(Player %in% c("Robert Covington","Thomas Bryant","Jordan Goodwin", "Kenrich Williams")) %>%
  mutate_at(vars(PTS, TRB, AST, STL, BLK, TOV, MP), ~./G) %>%
  select(c(Player, Pos, PTS, 'FG%', TRB, AST, STL, BLK, TOV, MP))
```

Looking at the box-score stats for these 4 players, we can see that they tended to average less points per game and assists per game than Caldwell-Pope, McDaniels, and Smart, but their rebounding numbers were fairly strong comparatively. The main difference between these 4 players and the previous 3 players I named is the disparity in the minutes played per game. In my mind, players who could be replaced by a replacement level player are those that struggle to find consistent minutes or roles on their team.

```{r}
NBA23 %>%
  group_by(Pos) %>%
  filter(n() > 2) %>%
  summarise(mean=mean(VORP)) %>% 
  arrange(desc(mean))
```

Once again, we can see that centers have the best average rating, while power forwards come in third. It is very interesting that most of these statistics seem to have centers as the best-rated.

```{r}
VORPmodel <- lm(VORP~Pos, data = NBA23)
summary(VORPmodel)
anova(VORPmodel)
```

From the analysis of variance, we can see that position has a significant association on VORP, with center, power forward, and point guard all being significantly higher in terms of average VORP as compared to shooting guards. This could indicate a possibly biased metric, similar to the other metrics we have looked at, as it is saying that some positions are more valuable than others. Perhaps this is not biased and it is simply the case that the center, point guard, and power forward positions are the most important in basketball or that those positions currently have the best players in the league.

## What does this tell us?

After looking into a few of the most popular "advanced" metrics that are used to evaluate a player's performance, value, and contribution, we can see that there may be some flaws with the current metrics. First off, each metric seemed to have some biases towards favoring bigs, especially centers. This could be due to the fact that the metrics put more importance on areas of basketball that centers normally thrive at, including rebounding, efficiency, low turnovers, etc. It has become clear that the superstar players will be near the top of each metric regardless of position because they put up great counting stats, but it is the non-star players I am interested in. Because bigs tend to have better metrics on average, this most likely occurs because the non-star centers are being rated better than the non-star wings and guards. We have also seen that counting stats (points, rebounds, assists, blocks, steals, turnovers) have a large impact on how players are rated, but this does not always tell the full story. Some players may be at a natural advantage to acquire these stats due to the role they play or position they are put in, but it does not make them inherently better or more valuable to their team.

Now, maybe centers are simply more important to basketball and the NBA is overflowing with talent at the position. However, as an avid NBA fan, I know that wings and perimeter players tend to be more coveted than bigs when it comes to free agency, trades, draft picks, and the buyout market. Every team is always looking for another versatile wing that can shoot, handle, play defense, etc. Additionally, a lot of teams prefer to go small-ball and they have players playing out of their normal position. Yet, shooting guards and small forwards consistently performed bad, on average, in the metrics we have investigated. I think one reason for this is due to how lineups are set and positions are identified now. Players like LeBron James and Kevin Durant are now considered power forwards even though they primarily played small forward for most of their career. Jimmy Butler, who was once primarily a shooting guard and small forward, is listed as a power forward.

Positions have become almost irrelevant, and due to the advanced skill sets of players in today's NBA, many players could realistically play multiple positions with ease. To show how positional distinctions do not tell us much about players, we can look at which stats are associated with each position.

## Which statistics are associated with each position?

```{r}
library(cluster)
library(factoextra)
NBA23 <- read.csv("~/Desktop/Basketball-Analytics/NBA23.csv",check.names=F)
head(NBA23)
```

I will use anova to identify which of the variables in the NBA data set are most closely associated to each position in basketball. If there are many similarities between positions, this could support the idea that position lines are blurred and they do not hold much value or tell us much about players anymore.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'PG'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
PG <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Point Guard\n")
print(PG)
```

The variables most significantly associated with the point guard position are assist percentage, percentage of their made 2 and 3 pointers which were assisted, number of assists, and total rebound percentage.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'SG'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
SG <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Shooting Guard\n")
print(SG)
```

The variables most significantly associated with the shooting guard position have to do with rebounding percentages, block percentage, 2 point percentage, and win shares.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'SF'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
SF <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Small Forward\n")
print(SF)
```

The variables most associated with the small forward position have to do with shot distances, assist percentage, turnover percentage, and percent of 2 point field goals assisted.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'PF'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
PF <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Power Forward\n")
print(PF)
```

The variables most significantly associated with the power forward position are rebounding percentages, percent of made shots that are assisted, dunks, and 2 point field goal percentage.

```{r}
# Specify the category you are interested in (e.g., 'PG')
category_of_interest <- 'C'

# List of numeric variable names
numeric_vars <- colnames(NBA23)[sapply(NBA23, is.numeric)]

# Initialize an empty list to store significant ANOVA results
significant_anova_results_list <- list()

# Iterate over each numeric variable
for (var in numeric_vars) {
  # Perform ANOVA comparing the category of interest with each other category
  anova_result <- aov(get(var) ~ factor(Pos == category_of_interest), data = NBA23)
  
  # Check if the p-value is significant
  if (summary(anova_result)[[1]][[5]][[1]][1] < 0.05) {
    # Store ANOVA result in the list
    significant_anova_results_list[[var]] <- summary(anova_result)
  }
}

significant_variables <- names(significant_anova_results_list)
p_values <- sapply(significant_anova_results_list, function(result) result[[1]][[5]][[1]][[1]])

# Create a data frame with significant variables and their p-values
significant_df <- data.frame(Variable = significant_variables, P_Value = p_values)

# Sort the data frame by p-values in ascending order
C <- significant_df[order(significant_df$P_Value), ]

# Display the table
cat("Center\n")
print(C)
```

Similar to the power forward position, the variables most significantly associated with the center position are rebounding percentages and dunks. Other important variables are block percentage, field goal percentage, and shot distance.

### Positional Similarities

```{r}
PG_SG <- intersect(PG$Variable, SG$Variable)
SG_SF <- intersect(SG$Variable, SF$Variable)
SF_PF <- intersect(SF$Variable, PF$Variable)
PF_C <- intersect(PF$Variable, C$Variable)

# Determine the maximum length among the vectors
max_length <- max(length(PG_SG), length(SG_SF), length(SF_PF), length(PF_C))

# Pad shorter vectors with NA to match the maximum length
PG_SG <- c(PG_SG, rep(NA, max_length - length(PG_SG)))
SG_SF <- c(SG_SF, rep(NA, max_length - length(SG_SF)))
SF_PF <- c(SF_PF, rep(NA, max_length - length(SF_PF)))
PF_C <- c(PF_C, rep(NA, max_length - length(PF_C)))

# Combine padded vectors into a data frame
similarities <- data.frame(PG_SG = PG_SG, SG_SF = SG_SF, SF_PF = SF_PF, PF_C = PF_C)

# Print the resulting data frame
print(similarities)
```

## My Proposed Solution

...

## 
